---
title: "ToxicRToo: Continuous Analysis"
author: "Matt Wheeler Ph.D."
date: "2/15/2022"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{ToxicRToo: Continuous Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)

# Detect optional dependencies used in this vignette
has_bioc <- all(vapply(c("SummarizedExperiment", "MultiAssayExperiment", "BiocParallel", "S4Vectors"),
                       requireNamespace, quietly = TRUE, FUN.VALUE = logical(1)))
has_gdr  <- all(vapply(c("gDRimport", "gDRcore"), requireNamespace, quietly = TRUE, FUN.VALUE = logical(1)))
```

## Abstract and Bioconductor Motivation

ToxicRToo provides Bayesian and frequentist dose–response modeling for continuous, binary, and count outcomes, with support for model averaging and rigorous uncertainty quantification. Although originally developed for environmental risk assessment, these methods are directly applicable to toxicogenomics, where we quantify how molecular features (e.g., gene expression) change with exposure. This vignette adds light-weight wrappers that interface ToxicRToo with common Bioconductor classes (SummarizedExperiment and MultiAssayExperiment) so that users can compute gene-level benchmark doses (BMD) and combine results with existing feature metadata.

Relation to existing Bioconductor packages:

- ToxicoGx: focuses on curated toxicogenomic datasets and dose–response summaries; ToxicRToo complements this by providing advanced single- and multi-model BMD estimation (frequentist and Bayesian) and model averaging.
- gDRimport/gDRcore: provide pipelines and infrastructure for high-throughput dose–response screens; ToxicRToo can serve as a modeling backend to derive BMD and interval estimates, with wrappers that accept SummarizedExperiment/MultiAssayExperiment used in those ecosystems.

## Interoperability with Bioconductor Classes (Wrappers)

This section demonstrates practical wrappers that accept SummarizedExperiment and MultiAssayExperiment so ToxicRToo can be used directly on common Bioconductor data containers. These examples use synthetic data for reproducibility in the vignette but apply equally to real datasets where `colData` contains a `dose` column and assays contain feature-level measurements.

### Setup

```{r bioc-setup, message=FALSE}
if (has_bioc) {
  library(SummarizedExperiment)
  library(MultiAssayExperiment)
  library(BiocParallel)
} else {
  message("Bioconductor packages not installed; Bioc wrappers/examples will be skipped.")
}
library(ToxicRToo)
```

### Wrapper: SummarizedExperiment -> ToxicRToo input

The function below aggregates replicates by dose for one feature (row) of a SummarizedExperiment and fits a continuous model, returning a ToxicRToo fit. It assumes `colData(se)$dose` exists and the selected assay contains numeric measurements where larger values reflect increased response (set `adverseDirection` accordingly if needed).

```{r se-wrapper, eval=has_bioc}
toxicr_fit_from_SE <- function(se, feature, assay = 1, dose_col = "dose",
                               model_type = "hill", distribution = "normal-ncv",
                               fit_type = "laplace", BMR_TYPE = "sd", BMR = 1.0,
                               adverseDirection = c("increase", "decrease"), ...) {
  adverseDirection <- match.arg(adverseDirection)
  stopifnot(dose_col %in% colnames(colData(se)))
  a <- assay(se, assay)
  stopifnot(feature %in% rownames(a))
  y <- as.numeric(a[feature, , drop = TRUE])
  d <- as.numeric(colData(se)[[dose_col]])
  df <- data.frame(dose = d, val = y)
  agg <- aggregate(val ~ dose, df, function(x) c(mean = mean(x), sd = stats::sd(x), n = length(x)))
  means <- sapply(agg$val, function(v) v[["mean"]])
  sds   <- sapply(agg$val, function(v) if (is.na(v[["sd"]])) 0 else v[["sd"]])
  ns    <- sapply(agg$val, function(v) v[["n"]])
  Y <- cbind(means, ns, sds)
  single_continuous_fit(agg$dose, Y,
    model_type = model_type, distribution = distribution,
    fit_type = fit_type, BMR_TYPE = BMR_TYPE, BMR = BMR, ...
  )
}
```

### Wrapper: BMD for all features in a SummarizedExperiment

Compute BMD and credible/confidence intervals for each feature and return a DataFrame that can be merged into `rowData`.

```{r se-bulk-wrapper, eval=has_bioc}
toxicr_bmd_by_feature <- function(se, assay = 1, dose_col = "dose",
                                  model_type = "hill", distribution = "normal-ncv",
                                  fit_type = "laplace", BMR_TYPE = "sd", BMR = 1.0,
                                  BPPARAM = SerialParam(), ...) {
  feats <- rownames(se)
  res <- bplapply(feats, function(f) {
    fit <- try(
      toxicr_fit_from_SE(se, f, assay = assay, dose_col = dose_col,
                         model_type = model_type, distribution = distribution,
                         fit_type = fit_type, BMR_TYPE = BMR_TYPE, BMR = BMR, ...),
      silent = TRUE
    )
    if (inherits(fit, "try-error")) return(data.frame(rowname = f, BMD = NA_real_, BMDL = NA_real_, BMDU = NA_real_))
    stats <- summary(fit)
    data.frame(rowname = f, BMD = stats$BMD[2], BMDL = stats$BMD[1], BMDU = stats$BMD[3])
  }, BPPARAM = BPPARAM)
  res <- do.call(rbind, res)
  S4Vectors::DataFrame(res[match(rownames(se), res$rowname), c("BMD", "BMDL", "BMDU")])
}
```

### Wrapper: MultiAssayExperiment

For studies with multiple assays, you can select a specific experiment and pass it to the previous functions. This mirrors how `gDRcore` and related packages organize data.

```{r mae-wrapper, eval=has_bioc}
toxicr_bmd_from_MAE <- function(mae, experiment, assay = 1, dose_col = "dose", ...) {
  stopifnot(experiment %in% names(experiments(mae)))
  se <- experiments(mae)[[experiment]]
  toxicr_bmd_by_feature(se, assay = assay, dose_col = dose_col, ...)
}
```

### Reproducible example with a synthetic SummarizedExperiment

Below we create a toy SummarizedExperiment with five doses and triplicate measurements for three "genes". We then compute gene-level BMD using the Hill model.

```{r synthetic-se-example, fig.height=4.5, fig.width=6, eval=has_bioc}
set.seed(123)
genes <- paste0("gene", 1:3)
doses <- rep(c(0, 1, 2, 4, 8), each = 3)
col_data <- S4Vectors::DataFrame(dose = doses)
mat <- sapply(setNames(genes, genes), function(g) {
  mu <- c(10, 10, 9, 7, 5) # synthetic decreasing trend
  unlist(lapply(mu, function(m) rnorm(3, m, sd = 1)))
})
mat <- t(mat)
rownames(mat) <- genes
se <- SummarizedExperiment::SummarizedExperiment(assays = list(measure = mat), colData = col_data)

# Compute BMD for each feature
res <- toxicr_bmd_by_feature(se, assay = 1, dose_col = "dose",
                             model_type = "hill", distribution = "normal-ncv",
                             fit_type = "laplace", BMR_TYPE = "sd", BMR = 1.0)
res
```

```{r install_packages, eval=FALSE}
# Installation is disabled in vignettes for checks. To run gDR examples locally:
# install.packages(c("gDRimport", "gDRcore"))
```

Load the packages:

```{r load_packages, eval=has_gdr}
library(gDRimport)
library(gDRcore)
```

### Step 2: Import and Preprocess Data

Use `gDRimport` to import dose-response data from a CSV file:

```{r import_data, eval=has_gdr}
# Import data using gDRimport
example_data <- gDRimport::import_data("path_to_data_file.csv")

# Display the first few rows of the data
print(head(example_data))

# Preprocess the data using gDRcore
processed_data <- gDRcore::process_data(example_data)

# Display the structure of the processed data
str(processed_data)
```

### Step 3: Analyze Data Using ToxicRToo

Once the data is processed, it can be used with ToxicRToo's modeling functions. Below is an example of fitting a Hill model:

```{r run_toxicr, eval=has_gdr}
# Fit a Hill model using the processed data
hill_fit <- single_continuous_fit(
  processed_data$Dose,
  cbind(processed_data$Mean, processed_data$N, processed_data$SD),
  model_type = "hill",
  distribution = "normal",
  fit_type = "mcmc"
)

# Display the fitted model
print(hill_fit@full_model)

# Plot the model fit
plot(hill_fit)
```

### Step 4: Benchmark Dose Analysis

Perform benchmark dose analysis using the fitted model:

```{r benchmark_analysis, eval=FALSE}
# Example: use summary(hill_fit) to obtain BMD, BMDL, BMDU
# stats <- summary(hill_fit)
# stats$BMD
```

### Step 5: Model Averaging

Run model averaging across multiple models:

```{r model_averaging, eval=has_gdr}
# Define priors for model averaging
prior_list <- list(
  create_continuous_prior(create_prior_list(normprior(0, 1, -100, 100)), "hill", "normal"),
  create_continuous_prior(create_prior_list(normprior(0, 1, -100, 100)), "exp-5", "normal")
)

# Perform model averaging
ma_fit <- ma_continuous_fit(
  processed_data$Dose,
  cbind(processed_data$Mean, processed_data$N, processed_data$SD),
  fit_type = "laplace",
  BMR_TYPE = "sd",
  BMR = 0.5,
  model_list = prior_list
)

# Plot the model averaging results
plot(ma_fit)
```

This example demonstrates the full workflow from importing data using Bioconductor packages to performing advanced dose-response modeling and analysis using ToxicRToo.
